Computing likelihoods under the SMC for a general class of ARGs.

### 2024-07-05 paper.tex
Change supplement to Appendix

% deal with supplementary items
\newcommand{\supplementarysection}{%
  \setcounter{figure}{0}% Reset figure counter
  \let\oldthefigure\thefigure% Capture figure numbering scheme
  \renewcommand{\thefigure}{S\oldthefigure}% Prefix figure number with S
  \section{Supplementary section}% Set supplementary section
}

% adjust inter-paragraph spacing - this one's for you, Gertjan
\setlength{\parskip}{0.5\baselineskip}

@@

The likelihood of ARG defined by the set of edges $E$ and nodes $N$
given parameters $\theta$ then is
\begin{equation}\label{eq:full-lik}
    \mathcal{L}(\mathcal{G}|r, \lambda)
    =
    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(c, p) \in E} A_e(r) B_e(r, \lambda) .
\end{equation}
One significant advantage of this formulation is that the computation time
is linear in the number of edges, and an algorithm can be devised to 
to efficiently keep track of $I_e(t)$ (see supplementary information 
\ref{par:algo}).
% Although the first two factors are easily expressed in terms of numbers of edges and nodes,
% we will see below
% that in practice it is convenient to absorb these terms into per-edge contributions.
We can also accommodate stepwise changes in the 
coalescence rate (or recombination rate)
through time by intersecting the internode intervals with the rate change
intervals; rate changes along the genome can be incorporated similarly.
Finally, we have only considered the topology and branch length information of any ARG.
Extending this expression by adding the likelihood of observing a set of mutations
given an ARG is given by
\begin{equation}\label{eq:lik-mut}
    \mathcal{L}(\mathcal{D}|\mathcal{G}, \mu)
    =
    \frac{1}{\mathcal{M}!} \prod_{(c, p) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}
\end{equation}
where $\mathcal{M}$ the total number of mutations and $m_e$
the number of mutations along edge $(c,p)$ \citep{mahmoudi_bayesian_2022}.

@@

\pagebreak

\supplementarysection
\section*{Supplementary Information}

% % % % % % % % % % % % % % % % % %
\subsection{Algorithm} \label{par:algo}

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/ts_algo_2rows.png}
\caption{Description of algorithm: Moving along the genome we transition from
local tree $T_1$ to the next, $T_2$ in three steps. 
Edges that do not persist in $T_2$ are
removed first. Edges starting at the left coordinate of $T_2$ are inserted.
After deletion or insertion of an edge the count vector $I$ is updated along the
corresponding internode intervals. Following an insertion, and prior to updating $I$,
the likelihood for that edge is computed as $I$ then represents the total number of
lineages the lineage associated with this new edge can
coalesce with. By moving along the genome in this way, each edge gets visited exactly once.}
\label{fig:algo}
\end{figure}

Here, we describe how to efficiently compute the likelihood under the SMC
from an ARG stored using the \emph{tree sequence} encoding. %  (see fig. \ref{fig:smc-unary}).
The tree sequence format encodes relationships using \emph{edges}
just as described above,
along with additional structure that
allows us to sequentially recover the local trees
along the genome very efficiently, and to reason about
the differences between those trees \citep{kelleher_efficient_2016, ralph_efficiently_2020}.
The order of the edge deletions and insertions required to move from one local tree $T_1$ to
the next, $T_2$ imposes the required strict total ordering on edges
(see definition \ref{def:coal})
that is needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
Indeed, following the removal of edges that do not persist in $T_2$,
we simply declare that each subsequently added edge
can only coalesce with the already-visited edges 
that make up the (partially) reconstructed local tree $T_2$.

This allows us to compute a likelihood in a single pass over all edges.
More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
the number of extant lineages in each internode interval (see fig \ref{fig:algo}).
When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along the internode intervals
spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
For each subsequently inserted edge, we first compute the likelihood contribution
of that edge by computing the integral
in equation \eqref{eq:full-lik} using $I$, and then update the counts in $I$.
We further keep track of
the last parent a child node was connected to to detect recombination events.

>>>>

% adjust inter-paragraph spacing - this one's for you, Gertjan
% \setlength{\parskip}{0.5\baselineskip}

@@

The likelihood of ARG defined by the set of edges $E$ and nodes $N$
given parameters $\theta$ then is
\begin{equation}\label{eq:full-lik}
    \mathcal{L}(\mathcal{G}|r, \lambda)
    =
    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(c, p) \in E} A_e(r) B_e(r, \lambda) .
\end{equation}
One significant advantage of this formulation is that the computation time
is linear in the number of edges, and an algorithm can be devised to 
to efficiently keep track of $I_e(t)$ (see Appendix).
% Although the first two factors are easily expressed in terms of numbers of edges and nodes,
% we will see below
% that in practice it is convenient to absorb these terms into per-edge contributions.
We can also accommodate stepwise changes in the 
coalescence rate (or recombination rate)
through time by intersecting the internode intervals with the rate change
intervals; rate changes along the genome can be incorporated similarly.
Finally, we have only considered the topology and branch length information of any ARG.
Extending this expression by adding the likelihood of observing a set of mutations
given an ARG is given by
\begin{equation}\label{eq:lik-mut}
    \mathcal{L}(\mathcal{D}|\mathcal{G}, \mu)
    =
    \frac{1}{\mathcal{M}!} \prod_{(c, p) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}
\end{equation}
where $\mathcal{M}$ the total number of mutations and $m_e$
the number of mutations along edge $(c,p)$ \citep{mahmoudi_bayesian_2022}.

@@

\setcounter{secnumdepth}{2} % Print out appendix section numbers

\section*{Appendix}
\appendix

\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}

% % % % % % % % % % % % % % % % % %
\subsection*{Algorithm} \label{par:algo}
In this section we describe how to efficiently compute the likelihood under the SMC
from an ARG stored using the \texttt{tskit} tree sequence encoding. 
The tree sequence format encodes relationships using \emph{edges},
along with additional information that
allows us to sequentially recover (and reason about 
their differences) 
efficiently~\citep{kelleher_efficient_2016, ralph_efficiently_2020}.
The order of the edge deletions and insertions required to move from one local tree $T_1$ to
the next, $T_2$ imposes the strict total ordering on edges
(see definition \eqref{def:coal})
that is needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
Indeed, following the removal of edges that do not persist in $T_2$,
we simply declare that each subsequently added edge
can only coalesce with the already-visited edges 
that make up the (partially) reconstructed local tree $T_2$.

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/ts_algo_2rows.png}
\caption{Likelihood computation algorithm. 
Moving along the genome we transition from
local tree $T_1$ to the next, $T_2$ in three steps. 
Edges that do not persist in $T_2$ are
removed first. Edges starting at the left coordinate of $T_2$ are inserted.
After deletion or insertion of an edge the count vector $I$ is updated along the
corresponding internode intervals. Following an insertion, and prior to updating $I$,
the likelihood for that edge is computed as $I$ then represents the total number of
lineages the lineage associated with this new edge can
coalesce with. By moving along the genome in this way, each edge gets visited exactly once.}
\label{fig:algo}
\end{figure}

This allows us to compute a likelihood in a single pass over all edges.
More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
the number of extant lineages in each internode interval (see Fig.~\ref{fig:algo}).
When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along 
the internode intervals
spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
For each subsequently inserted edge, we first compute the likelihood contribution
of that edge by computing the integral
in equation \eqref{eq:full-lik} using $I$, and then update the counts in $I$.
We further keep track of
the last parent a child node was connected to to detect recombination events.

[TODO another paragraph summarising what you actually do]
